Constructing a universe for the setoid model

### 2020-05-13 paper.tex
Write more

These extensionality principles are not theorems of intensional type theory as
usually defined [cite], although they are consistent with it [cite].

As a consequence, these principles are often added as axioms when formalizing
mathematics using proof-assistants based on ITT. This solution, however, is far
from optimal, as the introduction of axioms destroys the nice computational
properties of the system, particularly canonicity. Unfortunately, one can show
that whenever equality is defined according to the usual inductive definition by
ML, there is no hope to have a theory where both canonicity and funext/propext
coexist.

The first step is therefore to give up the idea of defining equality as a single
inductive definition, and adopt the approach according to which equality should
be defined specifically for each type former in the type theory. That is, every
type former $A$ should be introduced together with its own notion of equality
$A^\sim$, and the internal notion of identity in type theory should refer to
that relation.
%
If one does this, then funext can be obtained by simply defining equality on
function types to be point-wise equality. That is, $(A \to B)^\sim\ f\ g :\equiv
\forall (x : A) , B^\sim\ (f x)\ (g x)$.

* One possible answer: setoids

@@

* Setoid TT: MLTT + setoid constructs

@@

@@

%% The system that we will consider in this paper follows the same motivation of
%% OTT, of giving a syntactic presentation of the setoid model in [cite].

\section{Setoid Type Theory}

\subsection{Setoid model}

The setoid interpretation gives rise to a model of a type theory with functional
and propositional extensionality. It is thus a way to explain extensionality in
a type-theoretic and computationally adequate way \cite{setoid-model}.

\todo{say something more about this}

\todo{Say a bit about how one can turn the setoid model to a setoid translation
  (strict model and such...)}

\todo{This gives a translation/explanation/implementation of extensionality in
  terms of intensional TT}

\subsection{SeTT}

>>>>

These extensionality principles are not theorems of intensional type theory as
usually defined [cite], although they are consistent with it [cite].
%
As a consequence, these principles are often added as axioms when formalizing
mathematics using proof-assistants based on ITT. This solution, however, is far
from optimal, as the introduction of axioms destroys the nice computational
properties of the system, particularly canonicity. Unfortunately, one can show
that whenever equality is defined according to the usual inductive definition by
Martin-Loef, there is no hope to have a theory where both canonicity and
funext/propext coexist.

The first step out of this is to give up the idea of equality as a single
inductive definition. Instead, equality should be defined specifically for each
type former in the type theory. That is, every type former $A$ should be
introduced together with its own notion of equality $A^\sim$.
%
If one does this, then funext can be obtained by simply defining equality on
function types to be point-wise equality. That is, $(A \to B)^\sim\ f\ g :\equiv
\forall (x : A) , B^\sim\ (f x)\ (g x)$.

%% * One possible answer: setoids

@@

%% * Setoid TT: MLTT + setoid constructs

@@

... TODO ...

@@

The current formulation of SeTT is already a satisfactory (\todo{maybe not the
  right word: satisfactory to who? according to what metric?}) presentation of
the setoid model, in that it includes all the features of the semantics that we
wished to reflect in the syntax. It is also arguably expressive enough to
formalize a good chunk of set-level mathematics. However, it is also currently
lacking in a serious way, as the lack of universes seriously limits its
introspection capabilities, and consequently its strength as a foundational
system. Note that SeTT already includes a (univalent) universe of strict
propositions; hence what we are referring to is a universe of setoids.

The inclusion of a universe of setoids in SeTT poses several interesting
questions. First of all, since a universe of setoids is itself a type in SeTT,
is must also be a setoid. However, it certainly cannot be a univalent one. One
question is therefore what kind of equality to equip such universe.
%
The second question is how one can justify such universe in the setoid model.
In particular what principles, if any, should be added to the metatheory in
order to be able to define the universe.

This paper is an attempt to give an answer to the questions above.
...TODO...

\section{Setoid model}

%% The setoid interpretation gives rise to a model of a type theory with functional
%% and propositional extensionality. It is thus a way to explain extensionality in
%% a type-theoretic and computationally adequate way \cite{setoid-model}.

By \emph{setoid model} we intend a class of models of (intensional) type theory
where contexts/closed types are interpreted as setoids, i.e. sets with an
equivalence relation, and dependent types are interpreted as dependent/indexed
setoids.
%
A setoid model was first given by M. Hofmann in \cite{hofmann-ext-int} in order
to give a semantics to extensionality principles such as functional and
propositional extensionality. Here we consider a similar model construction, due
to T. Altenkirch \cite{setoid-model}.

The peculiarity of this model is that it is presented in a type theoretic and
intensional metatheory. The metatheory is also very simple, requiring only $\Pi$
types, $\Sigma$ types, and a universe of definitionally proof-irrelevant
propositions itself closed under such type formers.
%
Notably, the use of strict prop makes this model a strict one, namely one where
every equation in the model is a definitional equality in the metatheory.  The
model itself validates the extensionality principles mentioned above, like
funext, proext, and quotient types. As a consequence, this model construction
provides a way to explain extensionality, by giving a way to implement it from
stratch in terms of a base theory that is both very simple and intensional.

This idea of implementing extensionality via this model can be made both more
precise and more practical. Indeed, the strictness of this model implies that we
can turn this model into a syntactic translation, from which a syntactic
presentation of the model can be derived. (See next section, and [cite mpc'19])
\todo{rewrite this.}

\subsection{Setoid Type Theory}
